Lessons learned and paths forward for rabies dog vaccination in Madagascar: a case study of pilot vaccination campaigns in Moramanga District

### 2021-04-03 paper/mdpi_paper.tex
updating with final proofed files

%  LaTeX support: latex@mdpi.com
%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).

@@

% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'. The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

@@

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, aerospace, agriculture, agriengineering, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asc, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci , beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci , buildings, cancers, carbon , catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, children, cleantechnol, climate, clockssleep, cmd, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, dairy, data, dentistry, designs , diagnostics, diseases, diversity, drones, econometrics, economies, education, electrochem, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, futurephys, galaxies, games, gastrointestdisord, gels, genealogy, genes, geohazards, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, ijerph, ijfs, ijgi, ijms, ijns, ijtpp, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmse, jnt, jof, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, medicina, medicines, medsci, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, mti, nanomaterials, ncrna, neuroglia, nitrogen, notspecified, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, physics, plants, plasma, polymers, polysaccharides, preprints , proceedings, processes, proteomes, psych, publications, quantumrep, quaternary, qubs, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, test, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem, wevj

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by:
% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, expressionofconcern, extendedabstract, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
% supfile = supplementary materials

@@

%=================================================================
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{xx}
\issuenum{1}
\articlenumber{5}
\pubyear{2019}
\copyrightyear{2019}
%\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}
\updates{yes} % If there is an update available, un-comment this line

%% MDPI internal command: uncomment if new journal that already uses continuous page numbers
%\continuouspages{yes}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Lessons learned and paths forward for rabies dog vaccination in
Madagascar: a case study of pilot vaccination campaigns in Moramanga
District}

% Authors, for the paper (add full first names)
\Author{Caitlynn Filla$^{1, \dagger}$, Malavika Rajeev$^{2,
\dagger}$\href{https://orcid.org/0000-0001-6030-4995}{\orcidicon}, Zoavina
Randriana$^{3}$, Chantal Hanitriniana$^{4}$, Radoniaina R.
Rafaliarison$^{3}$, Glenn Edosoa$^{5}$, Mamitiana
Andriamananjara$^{6}$, Nivohanitra P. Razafindraibe$^{6}$, José
Nely$^{7}$, Angelique Ferreira$^{3, 8}$, Annie Yang$^{2,
\ddagger}$, Fenomanana Daniel$^{3}$, Tara A. Clarke$^{3, 9}$, Zachary
Farris$^{3, 10}$, Terry Stone$^{8}$, Jochem Lastdrager$^{8}$, Tsiky
Rajaonarivelo$^{3}$, Katie Hampson$^{11}$, C. Jessica E.
Metcalf$^{2}$, Kim Valenta$^{1, 3}$}

% Authors, for metadata in PDF
\AuthorNames{Caitlynn Filla, Malavika Rajeev, Zoavina Randriana, Chantal
Hanitriniana, Radoniaina R. Rafaliarison, Glenn Edosoa, Mamitiana
Andriamananjara, Nivohanitra P. Razafindraibe, José Nely, Angelique
Ferreira, Annie Yang, Fenomanana Daniel, Tara A. Clarke, Zachary
Farris, Terry Stone, Jochem Lastdrager, Tsiky Rajaonarivelo, Katie
Hampson, C. Jessica E. Metcalf, Kim Valenta}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad University of Florida, Department of Anthropology
Gainesville, Florida, USA; \\
$^{2}$ \quad Department of Ecology and Evolutionary Biology, Princeton
University Princeton, NJ, United States; \\
$^{3}$ \quad The Mad Dog Initiative Akanin'ny Veterinera, Akaikiniarivo,
Ambatobe, Antananarivo 101, Madagascar; \\
$^{4}$ \quad Mention Zoologie et Biodiversité Animale, Faculté des
Sciences Université d'Antananarivo, Antananarivo, Madagascar; \\
$^{5}$ \quad Chargé des Maladies Tropicales Négligées Organisation
mondiale de la Santé Madagascar, Antananarivo, Madagascar; \\
$^{6}$ \quad Direction des Services Vétérinaires Ministère chargé de
l'Agriculture et de l'Élevage, Antananarivo, Madagascar; \\
$^{7}$ \quad Service contre les Maladies Endémo-épidémiques et
Tropicales Négligées Ministère de la Santé Publique, Antananarivo,
Madagascar; \\
$^{8}$ \quad Travelling Animal Doctors 919 Crossan Rd, Newark, Delaware,
USA; \\
$^{9}$ \quad North Carolina State University, Department of Sociology
and Anthropology Raleigh, North Carolina, USA; \\
$^{10}$ \quad Appalachian State University, Department of Health and
Exercise Science Boone, North Carolina, USA; \\
$^{11}$ \quad Boyd Orr Centre for Population and Ecosystem Health
Institute of Biodiversity, Animal Health and Comparative Medicine
University of Glasgow, Glasgow, UK; \\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:mrajeev@princeton.edu}{\nolinkurl{mrajeev@princeton.edu}}}

@@

% The commands \thirdnote{} till \eighthnote{} are available for further notes

% Simple summary

@@

@@

@@

% Pandoc citation processing

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}

@@

Canine rabies results in an estimated 60,000 human deaths per year
globally \citep{hampson2015}. These deaths are entirely preventable:
prompt post-exposure prophylaxis of humans exposed to rabies is highly
effective at preventing death and mass dog vaccination can interrupt
transmission in domestic dogs and eventually lead to disease elimination
\citep{fooks2014}. The World Health Organization (WHO) and its partners
have set a goal to eliminate human deaths due to canine rabies by the
year 2030 (`ZeroBy30')\citep{abela-ridder2016}. Annual dog vaccination
campaigns achieving at least 70\% coverage are the recommended target
for controlling rabies in domestic dog populations
\citep{worldhealthorganization2013}. However, achieving this coverage
target in low and middle income countries where the burden of human
rabies is highest can be challenging due to economic, ecological,
sociocultural, and political barriers \citep{fahrion2017}.

@@

In Madagascar, canine rabies has been endemic for over a century, and
for most of that period, the Institut Pasteur de Madagascar has provided
post-exposure prophylaxis free-of-charge to bite patients in the country
\citep{reynes2011}. Currently, there are only 31 clinics where these
human vaccines are available, and there is minimal dog vaccination due
to high costs to owners and lack of vaccine availability
\citep{rajeev2018}. Recent studies have estimated a high burden of human
deaths, masked by weak surveillance across the country
\citep{rajeev2020, rajeev2018}.

The veterinary sector is largely private, with 204 veterinarians
employed in hybrid private/public employment as designated district
veterinarians. While dog vaccination is rare, livestock officers and
veterinarians work together to implement cattle vaccination campaigns
for anthrax on an annual basis as mandated by the government (owners are
charged a fee per animal for these vaccines which vary by location)
\citep{mondiale2017evaluation}. No routine mass dog vaccinations have
been conducted on the island, although a few pilot programs have begun
in recent years, largely implemented by NGO-government partnerships.

Here, we summarize lessons learned through the implementation of pilot
vaccination programs in the Moramanga District of Madagascar, where
previous work has shown high incidence of dog rabies cases and human
rabies exposures. In 2018 and 2019, we deployed two different
vaccination strategies. In 2018, we carried out a larger scale
volunteer-led pilot vaccination campaign in two communes (sub-district
level) using a static point strategy. In 2019/2020, we provided
vaccines, all necessary supplies for vaccine administration, and a
per-vaccine fee to the District Veterinarian to vaccinate animals as
part of routine services delivered continuously. We compare time, human
resources, costs, and coverage estimates between these campaigns, and
using a demographic and vaccination model we further explore different
vaccination strategies based on what we learned during implementation.

@@

The Moramanga District is located midway between the central highlands
and the east coast of Madagascar, at an average altitude of 936 m. It
comprises 21 communes, covering approximately 7150 km\textsuperscript{2}
with an approximate human population of 347,000\citep{instat2021}.
Previous work in the district has established a high burden of rabies
exposures and deaths despite the availability of post-exposure
prophylaxis at the district hospital \citep{rajeev2018}. While Moramanga
is relatively close to the capital city of Antananarivo
(\textasciitilde{} 3 hours by bus), within the district, travel times
between locations are highly variable, with much of the population
living in more rural areas with limited access to roads and
transportation \citep{rajeev2020}. Before 2018, there were limited
animal rabies vaccination services, with most animal vaccines available
in the urban commune of Moramanga Ville where owners were often charged
\textgreater{} 15,000 Ariary (\textasciitilde{} 4.28 USD) per vaccine
administered.

@@

In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal
Doctors (TAD)) organized a pilot vaccination campaign in collaboration
with the Department of Veterinary Services, and the Ministry of Public
Health in the District of Moramanga. This campaign focused on two
communes in Moramanga, Moramanga Ville (the district center) and
Andasibe (a rural commune surrounding Andasibe National Park), where
previously high incidence of probable rabies exposures (Moramanga Ville)
and a high burden of deaths (Andasibe) had been recorded
\citep{rajeev2018}.

The campaign was planned as a series of static point vaccination
stations covering 1 - 3 fokontany (i.e., sub-communes) per day. A week
before the campaign dates, the vaccination team informed the chief of
the fokontany about the campaign and provided posters advertising the
date of the vaccine and that it would be available at no cost to owners.
During the campaign, we used Rabisin (10 mL vials with 1 mL per dose,
Boehringer Ingelheim) to vaccinate both dogs and cats presented that
were at least 1 month old. As part of the campaign, owners were surveyed
by vaccinators about how many dogs and cats they owned in total (split
by \textgreater{} 1 year vs.~\textless{} 1 year in order to avoid
language ambiguities that might result in excluding puppies and
kittens), as well as if their dogs were free roaming (no restrictions on
movement by the owner/`mirenyreny', tied/`mifatora', or
fenced/`mifefy'). Vaccinations were delivered at no cost to owners, but
as animal vaccination is generally thought of as a paid service in
Madagascar, owners were asked how much in Ariary they would be willing
to pay to have one animal vaccinated against rabies. For each animal
vaccinated, we recorded the species (cat or dog), sex, approximate age
in years, and whether the animal had been previously vaccinated.

To assess coverage, post-vaccination coverage surveys were conducted
according to previously established
methodology\citep{sambo2017, gibson2015}. All animals vaccinated were
concurrently marked with a colored, non-toxic, livestock crayon along
the top of the head or back. Between 4 - 6 PM when dogs are most active
\citep{sambo2017} and on the same day as the campaign, in each
vaccination location two transect surveys were done. Each team (two
people per team) walked a transect for 1 hour starting in opposite
directions and accompanied by a local guide to ensure that walking paths
did not overlap. Teams recorded any marked and unmarked dogs observed,
their roaming status (whether roaming, inside a fence, or tied), and
their approximate age (greater or less than 1 year of age).

\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig1}

\hypertarget{figure-1.-photos-from-the-2018-campaign.-topleft-advertisement-for-the-campaign-posted-on-the-door-of-the-fokontany-office-as-the-campaign-starts-topright-a-dog-post-vaccination-marked-with-a-crayon-bottomleft-a-basket-of-puppies-brought-for-vaccination-bottomright-a-line-of-owners-and-dogs-waiting-for-vaccination.}{%
\paragraph{\texorpdfstring{\textbf{Figure 1. Photos from the 2018
campaign}. \emph{Topleft: advertisement for the campaign posted on the
door of the fokontany office as the campaign starts; Topright: a dog
post-vaccination marked with a crayon; Bottomleft: a basket of puppies
brought for vaccination; Bottomright: a line of owners and dogs waiting
for
vaccination.}}{Figure 1. Photos from the 2018 campaign. Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination.}}\label{figure-1.-photos-from-the-2018-campaign.-topleft-advertisement-for-the-campaign-posted-on-the-door-of-the-fokontany-office-as-the-campaign-starts-topright-a-dog-post-vaccination-marked-with-a-crayon-bottomleft-a-basket-of-puppies-brought-for-vaccination-bottomright-a-line-of-owners-and-dogs-waiting-for-vaccination.}}

@@

\hypertarget{data-and-ethics-statement}{%
\paragraph{Data and ethics statement}\label{data-and-ethics-statement}}

All data were analysed in R version 4.0.2 (2020-06-22) \citep{R2020},
largely using the \texttt{tidyverse} package suite \citep{wickham2019}.
Geospatial data were mapped using the sf \citep{sf} package. All data
and code are available at \url{https://github.com/mrajeev08/mora_vax}.
The vaccinations were part of a public health campaign and routine
veterinary service provisioning carried out by the local veterinary
officials, the NGOs involved, and in partnership with the Ministry of
Public Health and the Department of Veterinary Services at the national
level. MDI also maintains national research permits (MICET permit:
\#130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs.

>>>>

%  LaTeX support: latex@mdpi.com
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

@@

% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept". The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

%% https://tex.stackexchange.com/questions/161431/how-to-solve-longtable-is-not-in-1-column-mode-error
\usepackage{xtab}

\makeatletter
\def\longtable{\@ifnextchar[\newlongtable@i \newlongtable@ii}
\def\newlongtable@i[#1]{%
\renewcommand{\endhead}{\ignorespaces}
\xtabular[#1]}
\def\newlongtable@ii{%
\renewcommand{\endhead}{\ignorespaces}
\xtabular}
\def\endlongtable{\endxtabular}
\makeatother

@@

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, adolescents, aerospace, agriculture, agriengineering, agronomy, ai, algorithms, allergies, analytica, animals, antibiotics, antibodies, antioxidants, appliedchem, applmech, applmicrobiol, applnano, applsci, arts, asi, atmosphere, atoms, audiolres, automation, axioms, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomechanics, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, brainsci, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coatings, colloids, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, crops, cryptography, crystals, curroncol, cyber, dairy, data, dentistry, dermato, dermatopathology, designs, diabetology, diagnostics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, entropy, environments, environsciproc, epidemiologia, epigenomes, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forensicsci, forests, fractalfract, fuels, futureinternet, futuretransp, futurepharmacol, futurephys, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, hazardousmatters, healthcare, hearts, hemato, heritage, highthroughput, histories, horticulturae, humanities, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijtm, ijtpp, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jzbg, kidney, land, languages, laws, life, liquids, literature, livers, logistics, lubricants, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, metabolites, metals, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, minerals, mining, modelling, molbank, molecules, mps, mti, nanoenergyadv, nanomanufacturing, nanomaterials, ncrna, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, nri, nursrep, nutrients, obesities, oceans, ohbm, onco, oncopathology, optics, oral, organics, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photochem, photonics, physchem, physics, physiolsci, plants, plasma, pollutants, polymers, polysaccharides, proceedings, processes, prosthesis, proteomes, psych, psychiatryint, publications, quantumrep, quaternary, qubs, radiation, reactions, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, solids, sports, standards, stats, stresses, surfaces, surgeries, suschem, sustainability, symmetry, systems, taxonomy, technologies, telecom, textiles, thermo, tourismhosp, toxics, toxins, transplantology, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wevj, women, world

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by:
% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
% supfile = supplementary materials

@@

%=================================================================
% MDPI internal commands
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2021}
\copyrightyear{2020}
%\externaleditor{Academic Editor: Firstname Lastname} % For journal Automation, please change Academic Editor to "Communicated by"
\datereceived{}
\dateaccepted{}
\datepublished{}
\hreflink{https://doi.org/} % If needed use \linebreak
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%%%% If original paper need add "Retraction", please release the following command!!%%%%%%
%\retractiondate{Date Month Year} % For example,  13 October 2020
%\retractionnoticeyear{Year}
%\retractionnoticevolume{0}
%\retractionnoticeidnumber{0000}
%\retractionnoticedoi{10.3390/xxx}

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, soul, multirow, microtype, tikz, totcount, changepage, paracol, attrib, upgreek, cleveref, amsthm, hyphenat, natbib, hyperref, footmisc, url, geometry, newfloat, caption

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================

% Full title of the paper (Capitalized)
\Title{Lessons learned and paths forward for rabies dog vaccination in
Madagascar: a case study of pilot vaccination campaigns in Moramanga
District}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Lessons learned and paths forward for rabies dog
vaccination in Madagascar: a case study of pilot vaccination campaigns
in Moramanga District}

% Authors, for the paper (add full first names)
\Author{Caitlynn Filla$^{1, \dagger}$, Malavika Rajeev$^{2, \dagger,
*}$\href{https://orcid.org/0000-0001-6030-4995}{\orcidicon}, Zoavina
Randriana$^{3}$, Chantal Hanitriniana$^{4}$, Radoniaina R.
Rafaliarison$^{3}$, Glenn Torrencelli Edosoa$^{5}$, Mamitiana
Andriamananjara$^{6}$, Nivohanitra P. Razafindraibe$^{6}$, José
Nely$^{7}$, Angelique Ferreira$^{3, 8}$, Annie Yang$^{2,
\ddagger}$, Fenomanana Daniel$^{3}$, Tara A. Clarke$^{3, 9}$, Zachary
Farris$^{3, 10}$, Terry Stone$^{8}$, Jochem Lastdrager$^{8}$, Tsiky
Rajaonarivelo$^{3}$, Katie Hampson$^{11}$, C. Jessica E.
Metcalf$^{2}$, Kim Valenta$^{1, 3}$}

% Authors, for metadata in PDF
\AuthorNames{Caitlynn Filla, Malavika Rajeev, Zoavina Randriana, Chantal
Hanitriniana, Radoniaina R. Rafaliarison, Glenn Torrencelli
Edosoa, Mamitiana Andriamananjara, Nivohanitra P. Razafindraibe, José
Nely, Angelique Ferreira, Annie Yang, Fenomanana Daniel, Tara A.
Clarke, Zachary Farris, Terry Stone, Jochem Lastdrager, Tsiky
Rajaonarivelo, Katie Hampson, C. Jessica E. Metcalf, Kim Valenta}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad University of Florida, Department of Anthropology
Gainesville, Florida, 32611 USA; \\
$^{2}$ \quad Department of Ecology and Evolutionary Biology, Princeton
University Princeton, NJ, 08544 United States; \\
$^{3}$ \quad The Mad Dog Initiative Akanin'ny Veterinera, Akaikiniarivo,
Ambatobe, Antananarivo 101, Madagascar; \\
$^{4}$ \quad Mention Zoologie et Biodiversité Animale, Faculté des
Sciences Université d'Antananarivo, Antananarivo, Madagascar; \\
$^{5}$ \quad Chargé des Maladies Tropicales Négligées Organisation
mondiale de la Santé Madagascar, Antananarivo, Madagascar; \\
$^{6}$ \quad Direction des Services Vétérinaires Ministère chargé de
l'Agriculture et de l'Élevage, Antananarivo, Madagascar; \\
$^{7}$ \quad Service contre les Maladies Endémo-épidémiques et
Tropicales Négligées Ministère de la Santé Publique, Antananarivo,
Madagascar; \\
$^{8}$ \quad Travelling Animal Doctors 919 Crossan Rd, Newark, Delaware,
USA; \\
$^{9}$ \quad North Carolina State University, Department of Sociology
and Anthropology Raleigh, North Carolina, USA; \\
$^{10}$ \quad Appalachian State University, Department of Health and
Exercise Science Boone, North Carolina, USA; \\
$^{11}$ \quad Boyd Orr Centre for Population and Ecosystem Health
Institute of Biodiversity, Animal Health and Comparative Medicine
University of Glasgow, Glasgow, UK; \\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:mrajeev@princeton.edu}{\nolinkurl{mrajeev@princeton.edu}}}

@@

% The commands \thirdnote{} till \eighthnote{} are available for further notes
% Simple summary

@@

%\conference{} % An extended version of a conference paper

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Encyclopedia
%\encyclopediadef{Instead of the abstract}
%\entrylink{The Link to this entry published on the encyclopedia platform.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

Canine rabies results in an estimated 60,000 human deaths per year
globally \citep{hampson2015}. These deaths are entirely preventable:
prompt post-exposure prophylaxis of humans exposed to rabies is highly
effective at preventing death and mass dog vaccination can interrupt
transmission in domestic dogs and eventually lead to disease elimination
\citep{fooks2014}. The World Health Organization (WHO) and its partners
have set a goal to eliminate human deaths due to canine rabies by the
year 2030 (`ZeroBy30')\citep{abela-ridder2016}. Annual dog vaccination
campaigns achieving at least 70\% coverage are the recommended target
for controlling rabies in domestic dog populations
\citep{world2018expert}. However, achieving this coverage target in low
and middle income countries where the burden of human rabies is highest
can be challenging due to economic, ecological, sociocultural, and
political barriers \citep{fahrion2017}.

@@

In Madagascar, canine rabies has been endemic for over a century, and
for most of that period, the Institut Pasteur de Madagascar has provided
post-exposure prophylaxis free-of-charge to bite patients in the country
\citep{reynes2011}. Currently, there are only 31 clinics where these
human vaccines are available, with one clinic serving on average greater
than 700,000 persons and over 20,000 bite patients treated annually
\citep{rajeev2020}. There is minimal dog vaccination due to high costs
to owners and lack of vaccine availability \citep{rajeev2018}. Recent
studies have estimated a high burden of human rabies deaths
(approximately 1000 deaths annually), masked by weak surveillance across
the country \citep{rajeev2020, rajeev2018}.

The veterinary sector is largely private and practices are largely
limited to urban areas, but approximately 204 veterinarians are employed
in hybrid private/public employment as designated district veterinarians
by the national government. While dog vaccination is rare, livestock
officers and veterinarians work together to implement cattle vaccination
campaigns for anthrax on an annual basis as mandated by the government
(owners are charged a fee per animal for these vaccines which vary by
location) \citep{mondiale2017evaluation}. No routine mass dog
vaccinations have been conducted on the island, although a few pilot
programs have begun in recent years, largely implemented by
NGO-government partnerships.

Here, we summarize lessons learned through the implementation of pilot
vaccination programs in the Moramanga District of Madagascar, where
previous work has shown high incidence of dog rabies cases and human
rabies exposures. In 2018 and 2019, we deployed two different
vaccination strategies. In 2018, we carried out a larger scale
volunteer-led pilot vaccination campaign in two communes (sub-district
level) using a static point strategy where owners brought their animals
to a fixed location for vaccination. In 2019/2020, we provided vaccines,
all necessary supplies for vaccine administration, and a per-vaccine fee
to the District Veterinarian to vaccinate animals as part of continuous
vaccination strategy, where vaccines were delivered alongside routine
services provided by the veterinarian. We compare time, human resources,
costs, and coverage estimates between these campaigns, and using a
demographic and vaccination model we further explore different
vaccination strategies based on what we learned during implementation.

@@

The Moramanga District is located midway between the central highlands
and the east coast of Madagascar, at an average altitude of 936 m. It
comprises 21 communes, covering approximately 7150 km\textsuperscript{2}
with an approximate human population of 347,000\citep{instat2021}.
Previous work in the district has established a high burden of rabies
exposures (42 - 110 per 100,000 persons) and deaths (1 - 3 deaths per
100,000 persons) despite the availability of post-exposure prophylaxis
at the district hospital \citep{rajeev2018}. While Moramanga is
relatively close to the capital city of Antananarivo (\textasciitilde{}
3 hours by bus), within the district, travel times between locations are
highly variable, with much of the population living in more rural areas
with limited access to roads and transportation \citep{rajeev2020}.
Before 2018, there were limited animal rabies vaccination services, with
most animal vaccines available in the urban commune of Moramanga Ville
where owners were often charged \textgreater{} 15,000 Ariary
(\textasciitilde{} 4.28 USD) per vaccine administered.

@@

In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal
Doctors (TAD)) organized a pilot vaccination campaign in collaboration
with the Department of Veterinary Services, and the Ministry of Public
Health in the District of Moramanga (fig 2). This campaign focused on
two communes in Moramanga, Moramanga Ville (the district center) and
Andasibe (a rural commune surrounding Andasibe National Park), where
previously high incidence of probable rabies exposures (Moramanga Ville)
and a high burden of deaths (Andasibe) had been recorded
\citep{rajeev2018}.

The campaign was planned as a series of static point vaccination
stations covering 1 - 3 fokontany (i.e., sub-communes) per day. A week
before the campaign dates, the vaccination team informed the chief of
the fokontany about the campaign and provided fliers advertising the
date of the vaccine and that it would be available at no cost to owners
(fig 2). During the campaign, we used Rabisin (10 mL vials with 1 mL per
dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented
that were at least 1 month old based on current WHO recommendations for
endemic settings \citep{morters2015, world2018expert}. Rabisin has a
manufacturor stated duration of protection of 1 year given one dose, and
of an additional three years if an additional dose is given
approximately one year after the first dose. As part of the campaign,
owners were surveyed by vaccinators about how many dogs and cats they
owned in total (split by \textgreater{} 1 year vs.~\textless{} 1 year in
order to avoid language ambiguities that might result in excluding
puppies and kittens), as well as if their dogs were free roaming (no
restrictions on movement by the owner/`mirenyreny', tied/`mifatora', or
fenced/`mifefy'). Vaccinations were delivered at no cost to owners, but
as animal vaccination is generally thought of as a paid service in
Madagascar, owners were asked how much in Ariary they would be willing
to pay to have one animal vaccinated against rabies (after being
informed that the current vaccination was being free). For each animal
vaccinated, we recorded the species (cat or dog), sex, approximate age
in years as reported by the owner, and whether the animal had been
previously vaccinated.

To assess coverage, post-vaccination coverage surveys were conducted
according to previously established
methodology\citep{sambo2017, gibson2015}. All animals vaccinated were
concurrently marked with a colored, non-toxic, livestock crayon along
the top of the head or back. At the end of each campaign day between
1600-1800 hours, when dogs are most active \citep{sambo2017}, two
transect surveys were conducted on vaccination campaign days in each
vaccination location by two teams (consisting of two volunteers and one
local guide) for one hour on separate paths and in opposite directions.
Marked and unmarked dogs were recorded as well as their roaming status
(i.e., roaming, inside a fence, or tied), and their approximate age
(\textgreater1 year or \textless{} 1 year of age).

\begin{figure}
\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig1} \caption{Photos from the 2018 campaign. Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination. Photo credit: Jochem Lastdrager, Traveling Animal Doctors.}\label{fig:fig1}
\end{figure}

@@

\hypertarget{data-and-ethics-statement}{%
\subsection{Data and ethics statement}\label{data-and-ethics-statement}}
